---
title: "Lab 09: Algorithmic Bias"
author: "Anaelle Gackiere"
date: "03-20-2026"
output: github_document
---

## Load Packages and Data

First, let's load the necessary packages:

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(janitor)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r}
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )

# Take a look at the data
glimpse(compas)
```

# Part 1

### Exercise 1

What are the dimensions of the COMPAS dataset? (Hint: Use inline R code and functions like nrow and ncol to compose your answer.) What does each row in the dataset represent? What are the variables?

Our dataset has`r nrow(compas)` rows and `r ncol(compas)` columns. Each row represents one person (an observation). The variables are `r names(compas)`.

### Exercise 2

Based on the ID variable, there are `r n_distinct(compas$id)` unique defendants are in the dataset, while there are `r nrow(compas)` rows total. This is the same as the number of rows, but I'm not sure I'm convinced by this method.

So I'll group by name, and I find that each observation is still unique and matches the rows, even if the names are shared.

```{r unique}
# See if same name appears with diff IDs
compas %>%
  group_by(name) %>%
  summarise(n_ids = n_distinct(id)) %>%
  filter(n_ids > 1)


# some people might share the same name, so I'll double check with DOB
compas %>%
  group_by(name, dob) %>%
  summarise(n_ids = n_distinct(id)) %>%
  filter(n_ids > 1)
```

### Exercise 3

Let’s examine the distribution of the COMPAS risk scores (decile_score)!

The distribution is positively skewed.

```{r distribution}

ggplot(compas, aes(x = decile_score)) +
  geom_histogram(alpha = 0.5, fill = "plum") +
  theme_classic()

```

### Exercise 4

The distribution of defendants by race

```{r race-distribution}
ggplot(compas, aes(x = race)) +
  geom_bar(alpha = 0.5, fill = "plum") +
  theme_classic()

```

The distribution of defendants by sex

```{r sex-distribution}
ggplot(compas, aes(x = sex)) +
  geom_bar(alpha = 0.5, fill = "plum") +
  theme_classic()
```

The distribution of defendants by age category

```{r age-distribution}
ggplot(compas, aes(x = age_cat)) +
  geom_bar(alpha = 0.5, fill = "plum") +
  theme_classic()
```

Demographics Distribution

```{r demographics}

compas_long <- compas %>%
  select(race, sex, age_cat) %>%
  pivot_longer(cols = everything(), names_to = "demographic", values_to = "value")

ggplot(compas_long, aes(x = value, fill = demographic)) +
  geom_bar(alpha = 0.7) +
  facet_wrap(~ demographic, scales = "free") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of Defendants by Demographic Group",
       x = "", y = "Count")

```

# Part 2: Risk Scores and Recidivism

### Exercise 5

Create a visualization showing the relationship between risk scores (decile_score) and actual recidivism (two_year_recid). Do higher risk scores actually correspond to higher rates of recidivism

```{r recidivism}
ggplot(compas, aes(x = factor(two_year_recid), y = decile_score, fill = factor(two_year_recid))) +
  geom_boxplot(alpha = 0.7) +
  theme_classic() +
  labs(title = "Risk Scores by Recidivism Outcome",
       x = "Recidivism (0 = No, 1 = Yes)",
       y = "Risk Score",
       fill = "Actual Recidivism")
```

### Exercise 6

Calculate the overall accuracy of the COMPAS algorithm.

```{r accuracy-test}
compas <- compas %>%
  mutate(compas_classification = case_when(
    decile_score >= 7 & two_year_recid == 1 ~ "TP",
    decile_score <= 4 & two_year_recid == 0 ~ "TN",
    decile_score >= 7 & two_year_recid == 0 ~ "FP",
    decile_score <= 4 & two_year_recid == 1 ~ "FN",
    TRUE ~ NA_character_  # scores 5-6 are ambiguous, exclude them
  ))

# accuracy calculation
compas %>%
  filter(!is.na(compas_classification)) %>%
  summarise(
    TP = sum(compas_classification == "TP"),
    TN = sum(compas_classification == "TN"),
    FP = sum(compas_classification == "FP"),
    FN = sum(compas_classification == "FN"),
    accuracy = (TP + TN) / (TP + TN + FP + FN)
  )



```

### Exercise 7

Only 68.45% of its predictions are correct, which is very concerning.

# Part 3

### Exercise 8

Based on the distribution, I notice a concerning disparity in the risk scores between black and white defendants. The distribution of white defendants is very positively skewed (so low Risk Score for most), while the one of black defendants is uniform.

```{r score-disparities}

compas_black_white <- compas %>%
  filter(race %in% c("African-American", "Caucasian"))

ggplot(compas_black_white, aes(x = decile_score, fill = race)) +
  geom_histogram(alpha = 0.5, position = "dodge", bins = 10) +
  theme_classic() +
  labs(title = "Distribution of Risk Scores by Race",
       x = "Risk Score",
       y = "Count",
       fill = "Race")

```

### Exercise 9

There is a disparity in the percentage of black defendants classified as high risk (38.56%) compared to those who are white (17.07%).

```{r risk-disparity}

compas_black_white %>%
  group_by(race) %>%
  summarise(
    total = n(),
    high_risk = sum(decile_score >= 7),
    pct_high_risk = high_risk / total * 100
  )
```

### Exercise 10

```{r accuracy-byrace}

# To filter for specific conditions:
non_recidivists <- compas %>%
  filter(two_year_recid == 0)

# False Positive Rate by race
fpr_results <- non_recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  summarise(
    total_non_recidivists = n(),
    false_positives = sum(decile_score >= 7),
    FPR = false_positives / total_non_recidivists * 100
  )

# False Negative Rate by race
fnr_results <- compas %>%
  filter(race %in% c("African-American", "Caucasian"),
         two_year_recid == 1) %>%
  group_by(race) %>%
  summarise(
    total_recidivists = n(),
    false_negatives = sum(decile_score <= 4),
    FNR = false_negatives / total_recidivists * 100
  )

# print false positive
fpr_results

# print false negative
fnr_results 

```

### Exercise 11

There's a concerning disparity in the false negative rates by race (much higher for white defendants) and fasle positive rates by race (much higher for black defendants).

```{r race-disparity-visual}

# rbind didn't work since mismatch names so I used bind_rows instead and created a metric and value column (I'm sure there's an easier way but this worked)

fpr_plot <- fpr_results %>%
  select(race, value = FPR) %>%
  mutate(metric = "FPR")

fnr_plot <- fnr_results %>%
  select(race, value = FNR) %>%
  mutate(metric = "FNR")

bind_rows(fpr_plot, fnr_plot) %>%
  ggplot(aes(x = race, y = value, fill = race)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  facet_wrap(~ metric) +
  theme_classic() +
  labs(title = "False Positive and False Negative Rates by Race",
       x = "Race", y = "Rate (%)")
```

# Part 4: Understanding Sources of Bias

### Exercise 12

Based on the plot below, the algorithm weighs prior convictions differently for different racial groups, particularly for minoritized populations.

```{r priorcount-distribution}

ggplot(compas, aes(x = priors_count, y = decile_score, color = race)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic() +
  labs(title = "Prior Convictions vs Risk Score by Race",
       x = "Prior Convictions",
       y = "Risk Score",
       color = "Race")
```

### Exercise 13

There is not sufficient evidence to support Northpointe's claim that the algorithm is fair. Based on the plot below, the same score does not mean the same probability of recidivism across groups.

```{r calibration-check}

compas %>%
  group_by(race, decile_score) %>%
  summarise(
    total = n(),
    recidivated = sum(two_year_recid == 1),
    recid_rate = recidivated / total * 100
  ) %>%
  ggplot(aes(x = decile_score, y = recid_rate, color = race)) +
  geom_line() +
  geom_point() +
  theme_classic() +
  labs(title = "Calibration: Recidivism Rate by Risk Score and Race",
       x = "Risk Score (Decile)",
       y = "Actual Recidivism Rate (%)",
       color = "Race")

```

# Part 5: Designing Fairer Algorithms

### Exercise 14

If I was tasked with creating a fairer risk assessment algorithm, these are the changes I would make to address the disparities I have observed: 1. Prior convictions may be considered proxies for race since certain races have historically been overpoliced; it demonstrates biased policing. 2. Develop an algorithm that penalizes false positive rates somehow.

Overall, the data itself is biased, and it is difficult to create a fair algorithm with biased data.

### Exercise 15

Different definitions of fairness can sometimes be mathematically incompatible with each other. What trade-offs might be involved in designing a “fair” algorithm for criminal risk assessment?

Since base rates differ between racial groups here, it's not possible to have equal false positive and negative rates across groups and calibration at the same time. If one prioritizes calibration, then the same scores mean the same level of risk across groups. However, I think that within the group, individuals bear the cost of group-level base rate differences. If one prioritizes equal error rates, then it's fairer at the individual level but scores mean different things for different groups (and the prediction accuracy might drop). Prioritizing the reduction of false positives means fewer people are wrongly convicted, but you run the chance of not labeling someone who is actually a recidivist.

### Exercise 16

I think transparency is very important, in that defendants should be able to know how risk scores were calculated and that they should be able to challenge them in court. This might mean creating open source risk assessment tools. They also should not be allowed in actually sentencing someone and there should be some sort of time limit to them (something you did fifty years ago should not impact your risk score now).

There should be policies in place that also require workers of the judicial system to take courses on algorithmic bias so that they fully understand the type of technology used in court. It's also important to highlight what the data directly shows and not make inferences since the algorithm uses data about arrests and not crimes directly. Stretch Goals

### Exercise 17

Distribution of prior convictions by race.

```{r priorcounts-byrace}

# I wanted to use one geom density but there were to many fill variables so I used facet wrap instead
ggplot(compas, aes(x = priors_count, fill = race)) +
  geom_histogram(alpha = 0.7, bins = 10) +
  facet_wrap(~ race, scales = "free_y") +
  theme_classic() +
  labs(title = "Prior Convictions by Race",
       x = "Prior Convictions",
       y = "Count",
       fill = "Race")

# Some groups have very small sample sizes so let's check
compas %>% count(race) %>% arrange(n)

# this should look better
ggplot(compas %>% filter(!race %in% c("Native American", "Asian", "Other")), 
       aes(x = priors_count, fill = race)) +
  geom_histogram(alpha = 0.7, bins = 15) +
  facet_wrap(~ race, scales = "free_y") +
  theme_classic() +
  labs(title = "Distribution of Prior Convictions by Race",
       x = "Prior Convictions",
       y = "Count",
       fill = "Race")


```

Even though these distributions mainly appeared to all be positively skewed, it's important to look at the y-axis. There are far more black defendants in the dataset than any other group, which is an important disparity to keep in mind.

### Exercise 18

These variables do show different distributions across racial groups. Differences in type of degree charge may contribute to disparities in risk scores.

```{r priorcounts-chargetype}

# note M = misdemeanor, F = Felony
compas %>%
  filter(!race %in% c("Native American", "Asian")) %>%
  ggplot(aes(x = priors_count, fill = race)) +
  geom_histogram(alpha = 0.7, bins = 15) +
  facet_grid(race ~ c_charge_degree, scales = "free_y") +
  theme_classic() +
  labs(title = "Prior Convictions by Race and Charge Type",
       x = "Prior Convictions",
       y = "Count",
       fill = "Race")

## average is easier to compare; note I included NA and AA despite the small sample sizes here, so careful comparing those to others.
compas %>%
  group_by(race, c_charge_degree) %>%
  summarise(avg_priors = mean(priors_count)) %>%
  ggplot(aes(x = race, y = avg_priors, fill = c_charge_degree)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  theme_classic() +
  labs(title = "Average Prior Convictions by Race and Charge Type",
       x = "Race",
       y = "Average Prior Convictions",
       fill = "Charge Type")

```
