---
title: "Lab 09: Algorithmic Bias"
author: "Anaelle Gackiere"
date: "03-20-2026"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(janitor)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r}
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )

# Take a look at the data
glimpse(compas)
```

Part 1

### Exercise 1

What are the dimensions of the COMPAS dataset? (Hint: Use inline R code and functions like nrow and ncol to compose your answer.) What does each row in the dataset represent? What are the variables?

Our dataset has```r nrow(compas)``` rows and ```r ncol(compas)``` columns. 
Each row represents a certain variable. The variables are ```r names(compas)```.

### Exercise 2

### Exercise 3

### Exercise 4

Part 2

### Exercise 5

### Exercise 6

### Exercise 7

Part 3

### Exercise 8

### Exercise 9

### Exercise 10

## Additional Exercises  

Almost there! Keep building on your work and follow the same structure for any remaining exercises. Each exercise builds on the last, so take your time and make sure your code is working as expected.  
