---
title: "Lab 09: Algorithmic Bias"
author: "Anaelle Gackiere"
date: "03-20-2026"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(janitor)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r}
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )

# Take a look at the data
glimpse(compas)
```

Part 1

### Exercise 1

What are the dimensions of the COMPAS dataset? (Hint: Use inline R code and functions like nrow and ncol to compose your answer.) What does each row in the dataset represent? What are the variables?

Our dataset has```r nrow(compas)``` rows and ```r ncol(compas)``` columns. 
Each row represents one person (an observation). The variables are ```r names(compas)```.

### Exercise 2

Based on the ID variable, there are ```r n_distinct(compas$id)``` unique defendants are in the dataset, while there are ```r nrow(compas)``` rows total. This is the same as the number of rows, but I'm not sure I'm convinced by this method.

So I'll group by name, and I find that each observation is still unique and matches the rows, even if the names are shared.
```{r unique}
# See if same name appears with diff IDs
compas %>%
  group_by(name) %>%
  summarise(n_ids = n_distinct(id)) %>%
  filter(n_ids > 1)


# some people might share the same name, so I'll double check with DOB
compas %>%
  group_by(name, dob) %>%
  summarise(n_ids = n_distinct(id)) %>%
  filter(n_ids > 1)
```


### Exercise 3

Let’s examine the distribution of the COMPAS risk scores (decile_score)! 

The distribution is positively skewed. 

```{r distribution}

ggplot(compas, aes(x = decile_score)) +
  geom_histogram(alpha = 0.5, fill = "plum4") +
  theme_classic()

```


### Exercise 4

Part 2

### Exercise 5

### Exercise 6

### Exercise 7

Part 3

### Exercise 8

### Exercise 9

### Exercise 10

## Additional Exercises  

Almost there! Keep building on your work and follow the same structure for any remaining exercises. Each exercise builds on the last, so take your time and make sure your code is working as expected.  
