---
title: "Lab 09: Algorithmic Bias"
author: "Anaelle Gackiere"
date: "03-20-2026"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(janitor)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r}
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )

# Take a look at the data
glimpse(compas)
```

Part 1

### Exercise 1

What are the dimensions of the COMPAS dataset? (Hint: Use inline R code and functions like nrow and ncol to compose your answer.) What does each row in the dataset represent? What are the variables?

Our dataset has```r nrow(compas)``` rows and ```r ncol(compas)``` columns. 
Each row represents one person (an observation). The variables are ```r names(compas)```.

### Exercise 2

Based on the ID variable, there are ```r n_distinct(compas$id)``` unique defendants are in the dataset, while there are ```r nrow(compas)``` rows total. This is the same as the number of rows, but I'm not sure I'm convinced by this method.

So I'll group by name, and I find that each observation is still unique and matches the rows, even if the names are shared.
```{r unique}
# See if same name appears with diff IDs
compas %>%
  group_by(name) %>%
  summarise(n_ids = n_distinct(id)) %>%
  filter(n_ids > 1)


# some people might share the same name, so I'll double check with DOB
compas %>%
  group_by(name, dob) %>%
  summarise(n_ids = n_distinct(id)) %>%
  filter(n_ids > 1)
```


### Exercise 3

Let’s examine the distribution of the COMPAS risk scores (decile_score)! 

The distribution is positively skewed. 

```{r distribution}

ggplot(compas, aes(x = decile_score)) +
  geom_histogram(alpha = 0.5, fill = "plum") +
  theme_classic()

```


### Exercise 4
The distribution of defendants by race
```{r race-distribution}
ggplot(compas, aes(x = race)) +
  geom_bar(alpha = 0.5, fill = "plum") +
  theme_classic()

```

The distribution of defendants by sex

```{r sex-distribution}
ggplot(compas, aes(x = sex)) +
  geom_bar(alpha = 0.5, fill = "plum") +
  theme_classic()
```

The distribution of defendants by age category
```{r age-distribution}
ggplot(compas, aes(x = age_cat)) +
  geom_bar(alpha = 0.5, fill = "plum") +
  theme_classic()
```

Demographics Distribution
```{r demographics}

compas_long <- compas %>%
  select(race, sex, age_cat) %>%
  pivot_longer(cols = everything(), names_to = "demographic", values_to = "value")

ggplot(compas_long, aes(x = value, fill = demographic)) +
  geom_bar(alpha = 0.7) +
  facet_wrap(~ demographic, scales = "free") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of Defendants by Demographic Group",
       x = "", y = "Count")

```

Part 2: Risk Scores and Recidivism

### Exercise 5
Create a visualization showing the relationship between risk scores (decile_score) and actual recidivism (two_year_recid). Do higher risk scores actually correspond to higher rates of recidivism

```{r recidivism}
ggplot(compas, aes(x = factor(two_year_recid), y = decile_score, fill = factor(two_year_recid))) +
  geom_boxplot(alpha = 0.7) +
  theme_classic() +
  labs(title = "Risk Scores by Recidivism Outcome",
       x = "Recidivism (0 = No, 1 = Yes)",
       y = "Risk Score",
       fill = "Actual Recidivism")
```


### Exercise 6
Calculate the overall accuracy of the COMPAS algorithm.

```{r accuracy-test}
compas <- compas %>%
  mutate(compas_classification = case_when(
    decile_score >= 7 & two_year_recid == 1 ~ "TP",
    decile_score <= 4 & two_year_recid == 0 ~ "TN",
    decile_score >= 7 & two_year_recid == 0 ~ "FP",
    decile_score <= 4 & two_year_recid == 1 ~ "FN",
    TRUE ~ NA_character_  # scores 5-6 are ambiguous, exclude them
  ))

# accuracy calculation
compas %>%
  filter(!is.na(compas_classification)) %>%
  summarise(
    TP = sum(compas_classification == "TP"),
    TN = sum(compas_classification == "TN"),
    FP = sum(compas_classification == "FP"),
    FN = sum(compas_classification == "FN"),
    accuracy = (TP + TN) / (TP + TN + FP + FN)
  )



```

### Exercise 7

Only 68.45% of its predictions are correct, which is very concerning. 

Part 3

### Exercise 8

Based on the distribution, I notice a concerning disparity in the risk scores between black and white defendants. The distribution of white defendants is very positively skewed (so low Risk Score for most), while the one of black defendants is uniform.

```{r score-disparities}

compas %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  ggplot(aes(x = decile_score, fill = race)) +
  geom_histogram(alpha = 0.5, position = "dodge", bins = 10) +
  theme_classic() +
  labs(title = "Distribution of Risk Scores by Race",
       x = "Risk Score",
       y = "Count",
       fill = "Race")

```


### Exercise 9
There is a disparity in the percentage of black defendants classified as high risk (38.56%) compared to those who are white (17.07%). 
```{r risk-disparity}

compas %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  summarise(
    total = n(),
    high_risk = sum(decile_score >= 7),
    pct_high_risk = high_risk / total * 100
  )
```


### Exercise 10

## Additional Exercises  

Almost there! Keep building on your work and follow the same structure for any remaining exercises. Each exercise builds on the last, so take your time and make sure your code is working as expected.  
